---
title: "Why use randngen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{understanding-prngs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(randngen)
```


# Introduction

Pseudo-random number generators (PRNGs) power much of what goes on "behind the scenes" in statistical and cryptographic settings. In `R`, there is a pseudo-random number generator present which allows users to generate random variables from a variety of distributions. To check the default PRNG used by your `R` version you can run the following: 

```{r}
# Using version 4.2.2 (at time of writing)
# Returns the methods used for
# 1. The "default" random number generation, 
# 2. Normal variable generation 
# 3. Discrete uniform variable generation

RNGkind()
```


The PRNG which R utilizes can be specified by specifying a `.Random.seed` argument in the beginning of an R script or by adjusting the default `kind` (and `normal.kind` and `sample.kind`) arguments in the `RNGkind` or `set.seed` argument. 

While it is possible for one to utilize their own PRNG through use of user-supplied random number generation, such an approach is pedagogically complex. For users interested in working with and learning about random number generation, present tools available still leave a degree of complexity and/or mystery around how random numbers are generated from the seed(s) and parameters supplied to a given PRNG. The `randngen` package provides a suite of PRNGs to which aim to be easy to use and flexible for understanding the relevant maths and algorithms implemented by allowing users to specify all relevant parameters. 


# From PRNG to Uniform Random Variables (and beyond!)

```{r}
library(randngen)

# Function from here: https://stackoverflow.com/questions/5468280/scale-a-series-between-two-points
range01 <- function(x){(x-min(x))/(max(x)-min(x))}


randngen::lcg(seed = 1234, n=10000)|>
  range01()|>
  hist(main= "0-1 Scaled Uniform Random Variables with randgen::lcg()")
```

```{r}
set.seed(1234)
runif(10000)|>
  hist(main= "0-1 Uniform Random Variables with runif()")
```

## Applying the Inverse Probability Transform to Generate Random Variables

<!--
Define Inverse Probability Transform. Make POC with qnorm
--->
```{r}
library(randngen)

randngen::lcg(seed = 1234, n=10000)|>
  range01()|>
  qnorm()|>
  hist(main= "Normal(0,1) Random Variables using\nscaled randgen::lcg() values and qnorm()")
```

```{r}
set.seed(1234)
runif(10000)|>
  qnorm()|>
  hist(main= "Normal(0,1) Random Variables using runif() and qnorm()")
```

# Other applications

- Simulation studies: use the same PRNG and seed values for your research and not have to worry about 

# Appendix

## R's history with PRNGs 

__Disclaimer:__ _This section was copied from output from Gemeni. I have not found information on this collected in one individual place. However I was told by ChatGPT that it "[...] is well-documented in R’s official NEWS files, which detail changes and new features introduced in each R version ...". If this information is misleading or false, please [open an issue](https://github.com/benyamindsmith/randngen/issues/new) or submit a pull request with more accurate information._
## R's History with PRNGs

R's base distributions have steadily evolved their Pseudo-Random Number Generators (PRNGs) to prioritize longer periods, better equidistribution, and improved accuracy. Here is an accurate timeline of R's PRNG infrastructure.

### 1. Early Versions of R (Pre-1.7.0)
* **Default PRNG**: Marsaglia-Multicarry
* **Default Normal RNG**: Kinderman-Ramage
* **Context**: Early R relied heavily on George Marsaglia's multiply-with-carry approach, which offered a period of over 2^60. 
* **Legacy**: The original normal random number generator was later found to have approximation issues in the tails of the distribution. To ensure older scripts remain reproducible, R preserves this specific legacy algorithm under the name `"Buggy Kinderman-Ramage"`. Other uniform alternatives available at the time included **Wichmann-Hill** and **Super-Duper**.

### 2. R 1.7.0 Series (April 2003)
* **Default PRNG**: Mersenne Twister (MT19937)
* **Default Normal RNG**: Inversion
* **Context**: This major update shifted the default uniform PRNG to the **Mersenne Twister**, known for its massive cycle period of 2^19937 - 1 and strong equidistribution properties. 
* **Reproducibility**: To handle the breaking change in random streams, this release introduced the `RNGversion()` function, allowing users to explicitly request the algorithms used in older versions of R.

### 3. Introduction of L'Ecuyer-CMRG
* **Default PRNG**: Mersenne Twister (Maintained)
* **Context**: R introduced the **L'Ecuyer-CMRG** (MRG32k3a) algorithm to solve the problem of random number generation in parallelized code.
* **Use Case**: It provides multiple independent streams of random numbers, ensuring that parallel simulations do not suffer from overlapping random sequences or compromised statistical properties.

### 4. R 3.6.0 Series (April 2019)
* **Default PRNG**: Mersenne Twister (Maintained)
* **Context**: R fundamentally changed how discrete uniform random numbers are generated, directly impacting functions like `sample()`.
* **Technical Fix**: Previous versions used a modulo arithmetic method that introduced a slight bias in discrete sampling. R 3.6.0 replaced this with a rejection sampling method, ensuring perfectly uniform probabilities when drawing integers.

### 5. R 4.x Series (Current Behavior)
* **Default PRNG**: Mersenne Twister (Maintained)
* **Context**: Base R continues to rely on MT19937 as the default, with L'Ecuyer-CMRG recommended for multi-threading. 
* **Ecosystem Expansion**: Base R natively lacks modern, state-of-the-art generators like the `xoshiro` or `xoroshiro` families. Users requiring these high-performance algorithms must rely on external packages, such as `randngen`, to bridge the gap between base R's established defaults and cutting-edge PRNG research.

### Summary of Defaults and Capabilities

| R Version / Era | Default PRNG | Normal RNG | Key Updates & Notes |
|---|---|---|---|
| **Before R 1.7.0** | Marsaglia-Multicarry | Kinderman-Ramage | Legacy methods preserved via `"Buggy Kinderman-Ramage"` |
| **R 1.7.0 Series** | Mersenne Twister | Inversion | MT19937 becomes the standard; `RNGversion()` introduced |
| **Later Additions** | Mersenne Twister | Inversion | L'Ecuyer-CMRG added for parallel processing and independent streams |
| **R 3.6.0** | Mersenne Twister | Inversion | Modulo-based discrete sampling replaced with rejection sampling in `sample()` |
| **R 4.x (Current)** | Mersenne Twister | Inversion | Relies on external packages like `randngen` for `xoshiro` variants |


### Reproducibility Across Versions

R allows explicit control of the RNG type and seed via `RNGkind()` and `set.seed()`, respectively, ensuring reproducibility even as PRNG options expand.

# References

1. Random function - RDocumentation. https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/Random.

2. Morris, T. P., White, I. R. & Crowther, M. J. Using simulation studies to evaluate statistical methods. Statistics in Medicine 38, 2074–2102 (2019).

3. Generating distributions from random number generators. Cross Validated https://stats.stackexchange.com/questions/637706/generating-distributions-from-random-number-generators.

4. L’Ecuyer, P. & Simard, R. TestU01. ACM Transactions on Mathematical Software 33, 1–40 (2007).

5. R Core Team. R News Files. https://cran.r-project.org/doc/manuals/r-release/

