---
title: "Why use randngen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{understanding-prngs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(randngen)
```


# Introduction

Pseudo-random number generators (PRNGs) power much of what goes on "behind the scenes" in statistical and cryptographic settings. In `R`, there is a pseudo-random number generator present which allows users to generate random variables from a variety of distributions. To check the default PRNG used by your `R` version you can run the following: 

```{r}
# Using version 4.2.2 (at time of writing)
# Returns the methods used for
# 1. The "default" random number generation, 
# 2. Normal variable generation 
# 3. Discrete uniform variable generation

RNGkind()
```


The PRNG which R utilizes can be specified by specifying a `.Random.seed` argument in the beginning of an R script or by adjusting the default `kind` (and `normal.kind` and `sample.kind`) arguments in the `RNGkind` or `set.seed` argument. 

While it is possible for one to utilize their own PRNG through use of user-supplied random number generation, such an approach is pedagogically complex. For users interested in working with and learning about random number generation, present tools available still leave a degree of complexity and/or mystery around how random numbers are generated from the seed(s) and parameters supplied to a given PRNG. The `randngen` package provides a suite of PRNGs to which aim to be easy to use and flexible for understanding the relevant maths and algorithms implemented by allowing users to specify all relevant parameters. 


# From PRNG to Uniform Random Variables (and beyond!)

```{r}
library(randngen)

# Function from here: https://stackoverflow.com/questions/5468280/scale-a-series-between-two-points
range01 <- function(x){(x-min(x))/(max(x)-min(x))}


randngen::lcg(seed = 1234, n=10000)|>
  range01()|>
  hist(main= "0-1 Scaled Uniform Random Variables with randgen::lcg()")
```

```{r}
set.seed(1234)
runif(10000)|>
  hist(main= "0-1 Uniform Random Variables with runif()")
```

## Applying the Inverse Probability Transform to Generate Random Variables

<!--
Define Inverse Probability Transform. Make POC with qnorm
--->
```{r}
library(randngen)

randngen::lcg(seed = 1234, n=10000)|>
  range01()|>
  qnorm()|>
  hist(main= "Normal(0,1) Random Variables using\nscaled randgen::lcg() values and qnorm()")
```

```{r}
set.seed(1234)
runif(10000)|>
  qnorm()|>
  hist(main= "Normal(0,1) Random Variables using runif() and qnorm()")
```

# Other applications

- Simulation studies: use the same PRNG and seed values for your research and not have to worry about 

# Appendix

## R's history with PRNGs 

__Disclaimer:__ _This section was copied from output from ChatGPT. I have not found information on this collected in one individual place. However I was told by ChatGPT that it "[...] is well-documented in R’s official NEWS files, which detail changes and new features introduced in each R version ...". If this information is misleading or false, please [open an issue](https://github.com/benyamindsmith/randngen/issues/new) or submit a pull request with more accurate information._

R's base distributions have evolved in their use of Pseudo-Random Number Generators (PRNGs) over time to improve accuracy, speed, and security. Each R version may implement different PRNGs depending on updates, which can impact reproducibility for code across versions. Here’s a summary of the key PRNGs used across different base R versions and the changes introduced over time.

### 1. **Early Versions of R (< R 1.0.0)**
   - In the earliest versions of R, PRNGs were less standardized and primarily focused on providing simple random number generation for basic statistical analysis.
   - These versions used basic implementations of Linear Congruential Generators (LCGs), which are not very robust for high-quality randomness requirements but were adequate for basic applications.

### 2. **R 1.0.0 to R 2.0.0**
   - **PRNG**: Mersenne Twister (MT19937) and L'Ecuyer-CMRG
   - **Details**: R 1.0.0 introduced the widely adopted Mersenne Twister algorithm (MT19937), which remains one of the primary generators in many versions of R. It has a period of \(2^{19937} - 1\), making it suitable for many computational tasks due to its very high periodicity.
   - **Additional**: The L'Ecuyer-CMRG generator was also added as an option to allow for multiple independent streams of random numbers, which is useful for parallel processing.

### 3. **R 2.1.0 to R 2.14.0**
   - **PRNG**: Default remains Mersenne Twister, with support for L'Ecuyer-CMRG, Super-Duper, and Knuth-TAOCP-2002.
   - **Details**: During these versions, the L'Ecuyer-CMRG (multiple-recursive generator for parallel streams), Super-Duper, and Knuth’s TAOCP-2002 algorithms were added as user-selectable options. Super-Duper and Knuth-TAOCP-2002 are both generators with specific application suitability (e.g., Knuth's generator was well-suited for sampling due to its statistical robustness).

### 4. **R 2.15.0 to R 3.5.0**
   - **PRNG**: Default Mersenne Twister, updated L'Ecuyer-CMRG, and additional options including Marsaglia-Multicarry.
   - **Details**: These versions kept Mersenne Twister as the default generator but improved the L'Ecuyer-CMRG implementation. Marsaglia-Multicarry, a complementary-multiply-with-carry generator, was added as a lightweight, fast alternative.

### 5. **R 3.6.0**
   - **PRNG**: Change in sampling algorithm (not the PRNG itself).
   - **Details**: Starting with R 3.6.0, there was a change to the sampling algorithm used by functions like `sample()`. This update improved randomness in sampling without replacement by ensuring each permutation has equal probability, which impacted reproducibility of code relying on sampling in previous versions.
   - **Impact**: While this was not a change to the PRNGs directly, it affected random number generation behavior in R functions, requiring some users to set compatibility options for older scripts.

### 6. **R 4.0.0 Onwards**
   - **PRNG**: Default remains Mersenne Twister with support for 64-bit integers.
   - **Details**: The sampling mechanism and random number generation were further refined for 64-bit systems, improving the speed and quality of random number generation on modern hardware.
   - **Additional**: The `RNGkind()` function allows users to set RNG kinds explicitly for reproducibility, with settings such as `"Mersenne-Twister"`, `"L'Ecuyer-CMRG"`, and newer options like `"Xoroshiro128+"` for applications needing faster, non-cryptographic random numbers.

### 7. **R 4.2.0 and Later**
   - **PRNG**: Addition of Xoroshiro128+ as an optional generator.
   - **Details**: R 4.2.0 introduced Xoroshiro128+, a PRNG known for being fast and having a reasonable trade-off in terms of randomness quality for applications not requiring cryptographic security. It offers lower memory overhead compared to Mersenne Twister, making it ideal for simulation-heavy tasks.

### Summary Table

| **R Version**      | **Default PRNG**        | **Other Available PRNGs**                 | **Notes**                                                   |
|--------------------|-------------------------|-------------------------------------------|-------------------------------------------------------------|
| R < 1.0.0          | Basic LCG               | -                                         | Simple LCGs with limited randomness quality.                |
| R 1.0.0 - 2.0.0    | Mersenne Twister        | L'Ecuyer-CMRG                             | First use of Mersenne Twister as the default PRNG.          |
| R 2.1.0 - 2.14.0   | Mersenne Twister        | Super-Duper, Knuth-TAOCP-2002             | Additional PRNGs introduced for specific needs.             |
| R 2.15.0 - 3.5.0   | Mersenne Twister        | Marsaglia-Multicarry, updated L'Ecuyer-CMRG | Added Marsaglia-Multicarry for faster random numbers.      |
| R 3.6.0            | Mersenne Twister        | -                                         | Updated sampling algorithm for better random permutations.  |
| R 4.0.0 - 4.1.0    | Mersenne Twister        | L'Ecuyer-CMRG, 64-bit support             | 64-bit integer support for PRNGs.                           |
| R 4.2.0+           | Mersenne Twister        | Xoroshiro128+                             | Introduced Xoroshiro128+ for faster generation.             |

### Reproducibility Across Versions
R allows explicit control of the RNG type and seed via `RNGkind()` and `set.seed()`, respectively, ensuring reproducibility even as PRNG options expand.

# References

1. Random function - RDocumentation. https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/Random.

2. Morris, T. P., White, I. R. & Crowther, M. J. Using simulation studies to evaluate statistical methods. Statistics in Medicine 38, 2074–2102 (2019).

3. Generating distributions from random number generators. Cross Validated https://stats.stackexchange.com/questions/637706/generating-distributions-from-random-number-generators.

4. L’Ecuyer, P. & Simard, R. TestU01. ACM Transactions on Mathematical Software 33, 1–40 (2007).
